Inference using context-dependent models of mutation

### 2017-05-09 writeups/context-dependence.tex
added coverage results to paper

The usage of fast sparse matrix methods described above make likelihood computation fast enough
to use a Metropolis-Hasting-based Markov chain Monte Carlo method
to sample from the posterior distribution on parameters.
To test this,
we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.5$, $\beta=1.0$, and $\gamma=0.5$
for 1 unit of time, 
resulting in around 460,000 possible mutations
and differences at around 186,000 sites between initial and final sequence.
We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
(without constraining these to be equal)
and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
counted $(5,3,1)$ T-mers,
and used the mcmc package \citep{mcmcR} to run a random walk sampler for 10,000 steps
(judged sufficient by examination of diagnostic plots).
We repeated this procedure separately on 60 independently simulated sequences,
and show the resulting credible intervals in Figure \ref{fig:ising_coverage}.
Poseterior medians are within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
Credible intervals were somewhat too narrow, as the true value fell within 95\% credible intervals
for between 55\% and 90\% of the simulations;
which is likely due to remaining nonindependence at the window sizes used.

\begin{figure}
    \begin{center} 
        \includegraphics{writeup-plots/coverage_results}
    \end{center} 
    \caption{
        Credible intervals for the four parameters of the dynamic Ising model,
        relative to the true value,
        from 60 MCMC runs on independently simulated sequences of $10^6$ bases
        differing at around 18.6\% of the sites.
        \label{fig:ising_coverage}}
\end{figure}

>>>>

The usage of fast sparse matrix methods described above make likelihood computation fast enough
to use a Metropolis-Hasting-based Markov chain Monte Carlo method
to sample from the posterior distribution on parameters.
To test this,
we evolved $10^6$ sites from the Ising model (described above) with $\lambda=0.5$, $\beta=1.0$, and $\gamma=0.5$
for 1 unit of time, 
resulting in around 460,000 possible mutations
and differences at around 186,000 sites between initial and final sequence.
We then placed an Exponential prior with mean 1 on both transitions $+ \to -$ and $- \to +$
(without constraining these to be equal)
and half-Gaussian priors with mean 0 and scale 3 on both $\beta$ and $\gamma$,
counted $(5,3,1)$ T-mers,
and used the mcmc package \citep{mcmcR} to run a random walk sampler for 10,000 steps
(judged sufficient by examination of diagnostic plots).
We repeated this procedure separately on 60 independently simulated sequences,
and show the resulting credible intervals in Figure \ref{fig:ising_coverage}.
Poseterior medians are within half a percent of the truth for $+ \to -$ and $- \to +$ (corresponding to $\lambda$)
and the temperature ($\beta$), but showed around 2\% error for the magnetic field $\gamma$.
As seen in Table \ref{tab:ising_coverage}, 
credible intervals were somewhat too narrow for shorter T-mers,
but $(6,2,2)$ T-mers were long enough with sufficient overhang to produce well-calibrated posteriors.

\begin{table}[ht]
\centering
\begin{tabular}{rrrr}
  \hline
    & (4,1,1) & (5,1,1) & (6,2,2) \\ 
  \hline
  $\lambda_-$ & 0.61 & 0.63 & 0.96 \\ 
  $\lambda_+$ & 0.87 & 0.86 & 0.85 \\ 
  $\beta$     & 0.76 & 0.81 & 0.94 \\ 
  $\gamma$    & 0.69 & 0.76 & 0.85 \\ 
   \hline
\end{tabular}
    \caption{
        Posterior coverage of the four parameters of the dynamic Ising model at four different T-mer sizes.
        Shown are the proportion of runs in which the true value fell within the 95\% credible interval,
        for T-mers of length 4, 5, and 6 with overhangs of length 1, 1, and 2 respectively.
        The total number of runs was 54 for $(6,2,2)$ T-mers and 114 for the others.
        \label{tab:ising_coverage}
    }
\end{table}

\begin{figure}
    \begin{center} 
        \includegraphics{writeup-plots/coverage_results}
    \end{center} 
    \caption{
        Credible intervals for the four parameters of the dynamic Ising model,
        calculated using $(6,2,2)$ T-mers,
        from 54 MCMC runs on independently simulated sequences of $10^6$ bases
        differing at around 18.6\% of the sites.
        Shown are relative intervals, obtained by dividing by the true value.
        \label{fig:ising_coverage}}
\end{figure}
